---
title: "Anomaly Calculation and Data Aggregation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anomaly Calculation and Data Aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
library(er.helpers)
library(dplyr)
library(tibble)
```

## Overview

Environmental time series frequently contain missing values, and international
standards from organisations like the World Meteorological Organisation (WMO)
specify exactly how much missing data is acceptable before an aggregated value
should not be reported. `er.helpers` provides two complementary sets of tools
for these situations:

- **Aggregation with criteria** — compute a summary statistic (mean, min, max,
  sum) only when a dataset meets user-defined missing-data thresholds.
- **Annual anomaly** — express each year's value relative to a reference period
  mean, with the same missing-data safeguards applied to the baseline.

---

## Aggregating data with missing-value criteria

`aggregate_with_criteria()` returns the requested aggregate, or `NA` if either
the total number of missing values or the length of the longest consecutive run
of missing values exceeds the specified thresholds.

Thresholds can be expressed as **counts** (integers ≥ 1) or **proportions**
(0 < value ≤ 1):

```{r aggregate-basics}
# A month of daily temperatures with a few missing days
# randomize sequence of daily temperatures and missing values to demonstrate consecutive-missing criteria
set.seed(42)
daily_temp <- sample(c(15:25, NA, NA, NA), size = 30, replace = TRUE)

# Allow up to 3 missing days and no more than 2 consecutive missing days
aggregate_with_criteria(daily_temp, max_missing = 3, max_consecutive = 2)

# Tigthen the criteria: allow at most 2 missing days and no more than 1 consecutive missing day
aggregate_with_criteria(daily_temp, max_missing = 2, max_consecutive = 1)
```

### Convenience wrappers

`mean_with_criteria()`, `min_with_criteria()`, `max_with_criteria()`, and
`sum_with_criteria()` call `aggregate_with_criteria()` with the appropriate
`fun` argument pre-set:

```{r convenience-wrappers}
mean_with_criteria(daily_temp, max_missing = 3, max_consecutive = 2)
min_with_criteria(daily_temp,  max_missing = 3, max_consecutive = 2)
max_with_criteria(daily_temp,  max_missing = 3, max_consecutive = 2)
sum_with_criteria(daily_temp,  max_missing = 3, max_consecutive = 2)
```

### Applying criteria across groups with `dplyr`

A common workflow is to compute a monthly aggregate from daily data:

```{r dplyr-workflow}
set.seed(42)

daily_data <- tibble(
    date = seq.Date(as.Date("2020-01-01"),
    as.Date("2020-12-31"), by = "day"),
    temperature = runif(366, min = 5, max = 25)) |>
  # Introduce ~8 % random missing values
  mutate(temperature = ifelse(runif(n()) < 0.08, NA, temperature))

monthly_means <- daily_data |>
  mutate(year  = format(date, "%Y"),
         month = format(date, "%m")) |>
  group_by(year, month) |>
  summarise(mean_temp = mean_with_criteria(
    temperature, max_missing = 3, max_consecutive = 2),
    .groups = "drop")

monthly_means
```

Months where the missing-data criteria are exceeded will show `NA`, making it
easy to flag them downstream.

---

## Calculating annual anomalies

`calc_annual_anomaly()` subtracts the mean of a reference period from every
observation. The reference period mean is itself computed with
`mean_with_criteria()`, so the same missing-data safeguards apply.

```{r anomaly-setup}
set.seed(7)

# Simulate 50 years of annual mean temperatures
temp_series <- tibble(
    year = 1971:2020,
    temperature = 12 + cumsum(rnorm(50, mean = 0.03, sd = 0.4)))

# Define the WMO 1981–2010 climate normal as the reference period
reference_period <- c(1981, 2010)
```

```{r anomaly-calc}
temp_series <- temp_series |>
  mutate(anomaly = calc_annual_anomaly(temperature,
   year, period = reference_period))

head(temp_series)
```

### Visualising anomalies

```{r anomaly-plot, fig.width=7, fig.height=4}
library(ggplot2)

ggplot(temp_series, aes(x = year, y = anomaly, fill = anomaly > 0)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = c("#0D94A3", "#AE4E51")) +
  labs(title    = "Annual temperature anomaly",
       subtitle = paste0("Relative to ", reference_period[1], "\u2013",
                         reference_period[2], " mean"),
       x = NULL, y = "\u00b0C") +
  theme_minimal()
```

### Handling missing data in the reference period

If observations are missing inside the reference period, `calc_annual_anomaly()`
passes `max_missing` and `max_consecutive` to `mean_with_criteria()`. If the
reference mean cannot be calculated the function returns `NA` for every year
and issues a warning:

```{r anomaly-missing, warning=TRUE}
# Introduce missing values in the reference period
temp_with_gaps <- temp_series$temperature
temp_with_gaps[11:15] <- NA  # years 1981–1985 are missing

anomalies_strict <- calc_annual_anomaly(
    x = temp_with_gaps,
    year = temp_series$year,
    period = reference_period,
    max_missing = 0,   # zero tolerance for missing baseline data
    max_consecutive = 0)

head(anomalies_strict)

# Relax the constraint: allow up to 20 % missing in the reference period
anomalies_relaxed <- calc_annual_anomaly(
    x = temp_with_gaps,
    year = temp_series$year,
    period = reference_period,
    max_missing = 0.20,
    max_consecutive = NULL  # no consecutive-missing constraint
)

head(anomalies_relaxed)
```

---

## Summary of key arguments

| Argument | Description | Accepts |
|---|---|---|
| `max_missing` | Maximum allowable missing values | Count (integer) or proportion (0–1) |
| `max_consecutive` | Maximum allowable consecutive missing values | Count (integer) or proportion (0–1) |
| `fun` | Aggregation function (`aggregate_with_criteria` only) | Any function, e.g. `mean`, `sum` |
| `period` | Start and end year of the anomaly baseline | `c(start, end)` or `NULL` for full range |

Setting a threshold to `NULL` or `1` disables that constraint entirely.
