---
title: "Likelihood Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Likelihood Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(er.helpers)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

When communicating the results of statistical trend tests to a non-technical
audience, plain p-values are rarely helpful. `er.helpers` provides functions
to translate p-values into plain-language likelihood categories that align with
two established frameworks:

- **Statistics New Zealand** (`scale = "statsnz"`) — five categories with
  directional wording suited to environmental improvement/worsening language.
- **IPCC** (`scale = "ipcc"`) — nine categories aligned with Intergovernmental
  Panel on Climate Change reporting conventions.

The key functions are:

| Function | Purpose |
|---|---|
| `get_likelihood_category()` | Classify a p-value into a likelihood term |
| `order_likelihood_levels()` | Return a correctly ordered factor of likelihood terms |
| `simplify_likelihood_levels()` | Collapse detailed categories into broader ones |

The built-in datasets `statsnz_likelihood_scale` and `ipcc_likelihood_scale`
contain the interval boundaries used by each scale.

---

## The likelihood scales

```{r scales}
# Statistics NZ scale
statsnz_likelihood_scale

# IPCC scale
ipcc_likelihood_scale
```

---

## Classifying p-values

`get_likelihood_category()` maps a numeric probability to a category string.
The `term_type` argument controls the directional wording:

| `term_type` | Use case |
|---|---|
| `"worsening-improving"` (default) | High p → worsening (e.g. water quality metrics) |
| `"improving-worsening"` | High p → improving |
| `"increasing-decreasing"` | For climate variables with no value judgement |
| `"likely-unlikely"` | Neutral language |

```{r classify, warning=FALSE}
p_values <- seq(0, 1, length.out = 11)

# Water quality example: large p → worsening
get_likelihood_category(p_values, term_type = "worsening-improving")

# Climate variable: direction without value judgement
get_likelihood_category(p_values, term_type = "increasing-decreasing")

# Neutral language
get_likelihood_category(p_values, term_type = "likely-unlikely")
```

### IPCC scale

```{r ipcc}
get_likelihood_category(p_values, scale = "ipcc")
```

### Using percentage p-values

If your p-values are expressed as percentages (0–100) rather than probabilities
(0–1), use `p_is = "percentage"`:

```{r percentage}
p_percent <- p_values * 100
get_likelihood_category(p_percent, p_is = "percentage")
```

---

## Ordering factors for plotting

Likelihood categories returned by `get_likelihood_category()` are character
vectors. Use `order_likelihood_levels()` to convert them to an ordered factor
that displays correctly in plots:

```{r order}
categories <- get_likelihood_category(p_values, 
term_type = "increasing-decreasing")

# As unordered character — levels appear alphabetically
factor(categories) |> levels()

# Ordered factor — levels follow the scale from least to most likely
order_likelihood_levels(categories) |> levels()
```

### Complete trend-analysis workflow

```{r workflow}
set.seed(42)

# Simulate annual data for 10 indicator time series
indicator_df <- tibble(
  indicator = rep(paste0("I", 1:10), each = 30),
  year      = rep(1991:2020, times = 10),
  value     = rnorm(300, mean = rep(runif(10, -0.1, 0.1), each = 30),
                    sd   = 0.5)
)

# Run Mann-Kendall test for each indicator
trend_results <- indicator_df |>
  group_by(indicator) |>
  summarise(
    mann_kendall(value, alternative = "two.sided"),
    .groups = "drop"
  )

# Classify p-values and order the factor
trend_results <- trend_results |>
  mutate(
    likelihood = get_likelihood_category(p_value,
                                         term_type = "increasing-decreasing") |>
      order_likelihood_levels()
  )

trend_results |> select(indicator, p_value, likelihood)
```

```{r workflow-plot, fig.width=6, fig.height=4}
trend_counts <- trend_results |>
  count(likelihood)

ggplot(trend_counts, aes(x = likelihood, y = n, fill = likelihood)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = pal_snz_trend5) +
  labs(title = "Distribution of trend likelihoods",
       x = NULL, y = "Number of indicators") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Simplifying categories

When the full five-category scale is too granular for a summary, use
`simplify_likelihood_levels()` to collapse "Likely X" and "Very likely X"
into a single "Likely or very likely X" bucket:

```{r simplify, warning=FALSE}
detailed <- get_likelihood_category(p_values,
 term_type = "increasing-decreasing") |>
  order_likelihood_levels()

levels(detailed)

simplified <- simplify_likelihood_levels(detailed)
levels(simplified)
```

---

## Adding likelihood categories to a combined trend table

A typical reporting workflow chains trend testing, likelihood classification,
and ordering in a single `mutate()` call:

```{r full-table}
trend_results |>
  mutate(
    direction  = get_likelihood_category(p_value,
                                         term_type = "increasing-decreasing"),
    direction  = order_likelihood_levels(direction),
    simplified = simplify_likelihood_levels(direction)
  ) |>
  select(indicator, p_value, direction, simplified)
```
