---
title: "Colour Palettes and Scale Transformations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colour Palettes and Scale Transformations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(er.helpers)
library(ggplot2)
library(tibble)
library(dplyr)
```

## Overview

`er.helpers` provides two families of colour palettes tailored for
environmental reporting at Statistics New Zealand:

| Family | Prefix | Use case |
|---|---|---|
| Statistics NZ web | `pal_snz_*` | StatsNZ website publications |
| EA19 report | `pal_ea19_*` | Environment Aotearoa 2019 print report |
| Map / point | `pal_point_*` | Spatial maps and scatter plots |

Within each family there are palettes for different situations:

| Suffix | Purpose |
|---|---|
| *(none)* | Nominal (unordered) categorical variables |
| `_alpha2` | Comparing a current year to one past year |
| `_trend2` | Two-level ordinal trend (e.g. improving / worsening) |
| `_trend3` | Three-level ordinal trend (e.g. improving / indeterminate / worsening) |
| `_trend5` | Five-level ordinal trend |
| `_nztcs_c` | NZ Threat Classification System — categories |
| `_nztcs_s` | NZ Threat Classification System — sub-categories (SNZ only) |

An additional `signed_sqrt_trans()` scale transformation is useful for
visualising trend magnitudes that span both negative and positive values.

---

## Viewing palettes

Use `scales::show_col()` to display any palette:

```{r show-palettes, fig.width=8, fig.height=4}
scales::show_col(pal_snz, borders = NA)
```

```{r show-palettes-ea19, fig.width=8, fig.height=4}
scales::show_col(pal_ea19, borders = NA)
```

```{r show-palettes-point, fig.width=8, fig.height=4}
scales::show_col(pal_point_set1, borders = NA)
```

---

## Nominal palettes (unordered categories)

Use the base `pal_snz` / `pal_ea19` palettes with `scale_fill_manual()` or
`scale_colour_manual()` for unordered categorical variables:

```{r nominal, fig.width=6, fig.height=4}
set.seed(1)
indicator_data <- tibble(
  indicator = paste0("Indicator ", 1:7),
  value     = runif(7, 50, 150)
)

ggplot(indicator_data, aes(x = reorder(indicator, value), y = value,
                           fill = indicator)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = pal_snz) +
  coord_flip() +
  labs(title = "pal_snz — nominal palette", x = NULL, y = "Value") +
  theme_minimal()
```

---

## Trend palettes (ordinal categories)

Trend palettes are designed to pair with `order_likelihood_levels()` so that
colours align with the direction and strength of a trend.

### Two levels

```{r trend2, fig.width=5, fig.height=3}
trend2_df <- tibble(
  trend = factor(c("Improving", "Worsening")),
  n     = c(12, 8)
)

ggplot(trend2_df, aes(x = trend, y = n, fill = trend)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = pal_snz_trend2) +
  labs(title = "pal_snz_trend2", x = NULL, y = "Count") +
  theme_minimal()
```

### Five levels — full likelihood scale

```{r trend5, fig.width=6, fig.height=4, warning=FALSE}
set.seed(7)
p_values <- runif(50)

trend5_df <- tibble(
  p_value    = p_values,
  likelihood = get_likelihood_category(p_value, term_type = "increasing-decreasing") |>
    order_likelihood_levels()
) |>
  count(likelihood)

ggplot(trend5_df, aes(x = likelihood, y = n, fill = likelihood)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = pal_snz_trend5) +
  labs(title = "pal_snz_trend5 — five-level trend palette",
       x = NULL, y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Year-comparison palettes

`pal_snz_alpha2` and `pal_ea19_alpha2` use a tinted base colour for the
earlier year and the full base colour for the current year:

```{r alpha2, fig.width=5, fig.height=3}
comparison_df <- tibble(
  year  = factor(c(2019, 2023), levels = c(2019, 2023)),
  value = c(82, 91)
)

ggplot(comparison_df, aes(x = year, y = value, fill = year)) +
  geom_col(show.legend = FALSE, width = 0.5) +
  scale_fill_manual(values = pal_snz_alpha2) +
  labs(title = "pal_snz_alpha2 — year comparison palette",
       x = NULL, y = "Value") +
  theme_minimal()
```

---

## NZ Threat Classification System palettes

`pal_snz_nztcs_c` (four categories) and `pal_snz_nztcs_s` (nine
sub-categories) are named vectors, so they map automatically to factor levels:

```{r nztcs, fig.width=6, fig.height=3}
nztcs_df <- tibble(
  status = factor(names(pal_snz_nztcs_c),
                  levels = names(pal_snz_nztcs_c)),
  n      = c(45, 120, 18, 310)
)

ggplot(nztcs_df, aes(x = status, y = n, fill = status)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = pal_snz_nztcs_c) +
  labs(title = "pal_snz_nztcs_c — NZ Threat Classification",
       x = NULL, y = "Number of species") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

---

## Signed square-root scale transformation

When a continuous variable spans a wide range of both positive and negative
values, a standard linear scale compresses small values near zero. The
`signed_sqrt_trans()` transformation stretches both tails symmetrically:

```{r signed-sqrt, fig.width=7, fig.height=4}
set.seed(22)
slopes_df <- tibble(
  station = paste0("S", 1:20),
  slope   = c(rnorm(10, mean = -2, sd = 0.5),
              rnorm(10, mean =  8, sd = 3))
)

ggplot(slopes_df, aes(x = reorder(station, slope), y = slope,
                      fill = slope > 0)) +
  geom_col(show.legend = FALSE) +
  scale_y_continuous(trans = signed_sqrt_trans()) +
  scale_fill_manual(values = c("#0D94A3", "#AE4E51")) +
  labs(title = "Signed square-root y-axis",
       x = NULL, y = "Sen's slope (signed\u221a)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Quick reference

```{r quick-ref, echo=FALSE}
tibble::tribble(
  ~Object,            ~`Colours`, ~`Description`,
  "pal_snz",          9L, "Nominal — StatsNZ web",
  "pal_snz_alpha2",   2L, "Year comparison — StatsNZ web",
  "pal_snz_trend2",   2L, "Trend 2-level — StatsNZ web",
  "pal_snz_trend3",   3L, "Trend 3-level — StatsNZ web",
  "pal_snz_trend5",   5L, "Trend 5-level — StatsNZ web",
  "pal_snz_nztcs_c",  4L, "NZ TCS categories — StatsNZ web",
  "pal_snz_nztcs_s",  9L, "NZ TCS sub-categories — StatsNZ web",
  "pal_ea19",         9L, "Nominal — EA19 report",
  "pal_ea19_alpha2",  2L, "Year comparison — EA19 report",
  "pal_ea19_trend2",  2L, "Trend 2-level — EA19 report",
  "pal_ea19_trend3",  3L, "Trend 3-level — EA19 report",
  "pal_ea19_trend5",  5L, "Trend 5-level — EA19 report",
  "pal_ea19_nztcs_c", 4L, "NZ TCS categories — EA19 report",
  "pal_point_set1",   9L, "Nominal — maps/points",
  "pal_point_trend2", 2L, "Trend 2-level — maps/points",
  "pal_point_trend3", 3L, "Trend 3-level — maps/points",
  "pal_point_trend5", 5L, "Trend 5-level — maps/points"
)
```
