---
title: "Precipitation Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Precipitation Indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(er.helpers)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

Heavy rainfall events are often characterised by the proportion of total annual
rainfall that falls on the most extreme days. `er.helpers` provides two
functions to calculate this index:

| Function | Purpose |
|---|---|
| `get_reference_precipitation()` | Compute a percentile threshold from a reference period |
| `precipitation_above_reference()` | Compute the proportion of annual rainfall above that threshold |

Together they implement the precipitation index commonly reported by the
Intergovernmental Panel on Climate Change (IPCC) and the Expert Team on Climate
Change Detection and Indices (ETCCDI).

---

## Core concepts

**Reference precipitation** — typically the 95th or 99th percentile of daily
rainfall on wet days during a climate normal (e.g. 1981–2010). Only days
where precipitation ≥ `wet_day_threshold` (default 1 mm) are included.

**Proportion above reference** — the fraction of total annual wet-day
rainfall that falls on days exceeding the reference threshold. An increase in
this proportion over time indicates that intense rainfall events are becoming
more dominant.

---

## Step 1 — Compute the reference threshold

`get_reference_precipitation()` returns a single percentile value calculated
from the rainfall data within a specified reference period:

```{r reference}
set.seed(101)

# Simulate 20 years of daily rainfall (log-normal, approx. 2 mm/day)
n_days      <- 20 * 365
rainfall    <- rlnorm(n_days, meanlog = 0.5, sdlog = 1.0)
years       <- rep(2001:2020, each = 365)

rain_data <- tibble(rainfall = rainfall, year = years)

# Climate normal: 2001–2010
climate_normal <- c(2001, 2010)

# 95th percentile of wet days during the climate normal
ref_95 <- get_reference_precipitation(
  precipitation = rain_data$rainfall,
  date = rain_data$year,
  reference_period = climate_normal,
  percentile = 95)

ref_95
```

When `reference_period = NULL` the entire series is used as the reference.
You can also request a different percentile, such as the 99th:

```{r reference-99}
ref_99 <- get_reference_precipitation(
  precipitation = rain_data$rainfall,
  date = rain_data$year,
  reference_period = climate_normal,
  percentile = 99
)

ref_99
```

### Using date objects as the period

The `date` and `reference_period` arguments can be any comparable type. Numeric
years (as above), `Date`, or `POSIXct` objects all work — just ensure both are
the same type:

```{r reference-dates, eval=FALSE}
rain_data_dates <- rain_data |>
  mutate(date = seq.Date(as.Date("2001-01-01"),
  by = "day",
  length.out = n()))

ref_dates <- get_reference_precipitation(
  precipitation = rain_data_dates$rainfall,
  date = rain_data_dates$date,
  reference_period = as.Date(c("2001-01-01", "2010-12-31"))
)
```

---

## Step 2 — Compute the annual proportion above reference

After calculating the reference threshold, use `precipitation_above_reference()`
within a grouped summary to obtain the annual index:

```{r proportion}
annual_index <- rain_data |>
  mutate(ref = get_reference_precipitation(
  rainfall, year, climate_normal, percentile = 95)) |>
  group_by(year) |>
  summarise(
    prop_above = precipitation_above_reference(rainfall, ref),
    .groups    = "drop")

annual_index
```

---

## Step 3 — Visualise and assess trends

```{r plot, fig.width=7, fig.height=4}
ggplot(annual_index, aes(x = year, y = prop_above)) +
  geom_line(colour = "#085c75") +
  geom_point(colour = "#085c75") +
  geom_smooth(method = "lm", se = FALSE, colour = "#AE4E51",
              linewidth = 0.7, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title    = "Proportion of annual rainfall from very wet days",
       subtitle = "95th percentile threshold from 2001\u20132010 climate normal",
       x = NULL, y = "Proportion of total rainfall") +
  theme_minimal()
```

Combine with the trend functions from the `trends` vignette to test whether
the index is changing significantly over time:

```{r trend-test}
annual_index |>
  summarise(
    mann_kendall(prop_above, alternative = "greater"),
    .groups = "drop"
  )
```

---

## Adjusting the wet-day threshold

By default, a wet day is defined as any day receiving ≥ 1 mm. You can change
this to match a different standard:

```{r wet-day}
# Different wet-day thresholds
propn_1mm  <- rain_data |>
  group_by(year) |>
  summarise(
    prop = precipitation_above_reference(
      rainfall,
      get_reference_precipitation(rainfall, year, climate_normal),
      wet_day_threshold = 1),
    .groups = "drop")

propn_0mm  <- rain_data |>
  group_by(year) |>
  summarise(
    prop = precipitation_above_reference(
      rainfall,
      get_reference_precipitation(rainfall, year, climate_normal,
      wet_day_threshold = 0),
      wet_day_threshold = 0),
    .groups = "drop")
```

Set `wet_day_threshold = 0` to include all days (including dry days) in the
denominator — equivalent to treating all days as wet.

---

## Summary of key arguments

| Argument | Default | Description |
|---|---|---|
| `precipitation` | — | Numeric vector of daily rainfall values |
| `date` | `NULL` | Corresponding dates or years |
| `reference_period` | `NULL` | Two-element vector defining the climate normal |
| `percentile` | `95L` | Threshold percentile (0–100) |
| `wet_day_threshold` | `1` | Minimum rainfall to count a day as wet (mm) |
