---
title: "Trend Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trend Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(er.helpers)
library(dplyr)
library(tibble)
library(ggplot2)
```

## Overview

`er.helpers` provides three tidy wrappers for common environmental trend tests.
Each function returns a one-row `tibble` so results integrate naturally with
`group_by()` / `summarise()` workflows.

| Function | Method | Best for |
|---|---|---|
| `mann_kendall()` | Mann-Kendall test | Non-parametric monotonic trend detection |
| `sen_slope()` | Sen's slope | Non-parametric slope estimation with CI |
| `linear_model()` | OLS linear regression | Parametric slope with $R^2$ |

All three:
- Accept a numeric vector `x` and an optional time variable `y`.
- Return `NA` results and a warning when `x` contains missing values.
- Include a `note` column that records non-fatal conditions (e.g. < 10
  observations).

---

## Simulating example data

```{r sim-data}
set.seed(42)
n <- 40

# A series with a real upward trend
trend_data <- tibble(
  year  = 1981:(1980 + n),
  value = 10 + 0.15 * (1:n) + rnorm(n, sd = 1.2)
)

# A flat series (no trend)
flat_data <- tibble(
  year  = 1981:(1980 + n),
  value = 10 + rnorm(n, sd = 1.2)
)
```

---

## Mann-Kendall trend test

The Mann-Kendall test is non-parametric and robust to non-normal distributions
and outliers. It tests the null hypothesis of **no monotonic trend**.

```{r mann-kendall}
mk_result <- mann_kendall(trend_data$value, alternative = "greater")
mk_result
```

Key columns in the output:

| Column | Description |
|---|---|
| `p_value` | p-value for the chosen alternative |
| `tau` | Kendall's τ (strength and direction of trend) |
| `s` | Mann-Kendall S statistic |
| `z` | Standardised test statistic |
| `alternative` | The hypothesis tested |
| `note` | Any non-fatal diagnostic messages |

The `alternative` argument follows the same convention as `trend::mk.test`:

- `"greater"` — test for an upward trend (most common for worsening indicators).
- `"less"` — test for a downward trend.
- `"two.sided"` — test for any monotonic trend.

```{r mk-flat}
# No trend expected
mann_kendall(flat_data$value, alternative = "two.sided")
```

---

## Sen's slope

Sen's slope is the median slope across all pairs of observations. It is more
robust to outliers than OLS regression and is typically reported alongside
Mann-Kendall:

```{r sen-slope}
ss_result <- sen_slope(trend_data$value)
ss_result
```

Key columns:

| Column | Description |
|---|---|
| `slope` | Sen's slope estimate (change per time step) |
| `conf_low` / `conf_high` | Confidence interval for the slope |
| `conf_level` | Confidence level used |
| `p_value` | p-value from the associated Kendall test |

---

## Linear model

`linear_model()` fits OLS regression and returns a tidy summary including the
slope, confidence interval, $R^2$, and residual standard error. It optionally
accepts a `y` argument for the time axis:

```{r linear-model}
lm_result <- linear_model(trend_data$value, y = trend_data$year)
lm_result
```

When `y` is omitted, observations are assumed to be equally spaced (index 1, 2,
3, …). Passing `Date` or `POSIXct` vectors is also supported:

```{r linear-model-dates}
date_series <- tibble(
  date  = seq.Date(as.Date("2000-01-01"), by = "year", length.out = 20),
  value = cumsum(rnorm(20, mean = 0.5))
)

linear_model(date_series$value, y = date_series$date)
```

---

## Running multiple trend tests at once

The one-row-tibble design makes it easy to apply all three methods to many
groups with `group_by()` + `summarise()`:

```{r multi-group}
set.seed(99)

many_series <- tibble(
  region = rep(c("North", "South", "East", "West"), each = 30),
  year   = rep(2000:2029, times = 4),
  value  = c(
    5  + 0.20 * (1:30) + rnorm(30),   # North: increasing
    8  - 0.10 * (1:30) + rnorm(30),   # South: decreasing
    12 + rnorm(30),                    # East: flat
    6  + 0.05 * (1:30) + rnorm(30)    # West: weak increase
  )
)

trend_summary <- many_series |>
  group_by(region) |>
  summarise(
    sen_slope(value),
    .groups = "drop"
  ) |>
  mutate(
    likelihood = get_likelihood_category(p_value, term_type = "increasing-decreasing") |>
      order_likelihood_levels()
  )

trend_summary |> select(region, p_value, slope, conf_low, conf_high, likelihood)
```

---

## Handling missing data

All three functions return `NA` results with a warning when the input contains
`NA` values; the `note` column records this:

```{r missing}
x_with_na <- trend_data$value
x_with_na[c(5, 10, 15)] <- NA

mann_kendall(x_with_na)
```

Remove missing values before calling the trend functions, or impute them using
the aggregation tools in the `anomaly-and-aggregation` vignette.

---

## Comparing all three methods

```{r compare, fig.width=7, fig.height=4}
results <- bind_rows(
  mann_kendall(trend_data$value) |> mutate(slope = tau) |> select(p_value, slope, method),
  sen_slope(trend_data$value)    |> select(p_value, slope, method),
  linear_model(trend_data$value) |> select(p_value, slope, method)
)

results

# Plot the trend line from the linear model
lm_fit <- linear_model(trend_data$value, y = trend_data$year)

ggplot(trend_data, aes(x = year, y = value)) +
  geom_point(colour = "#085c75") +
  geom_abline(intercept = lm_fit$intercept,
              slope     = lm_fit$slope,
              colour    = "#AE4E51", linewidth = 0.8) +
  labs(title    = "Annual values with fitted linear trend",
       subtitle = paste0("slope = ", round(lm_fit$slope, 3),
                         ", p = ", signif(lm_fit$p_value, 3)),
       x = NULL, y = "Value") +
  theme_minimal()
```
